Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-10 paper.tex
update ARG recording section

Formulated backwards-in-time the SMC \citep{mcvean_approximating_2005} only requires a simple modification to the 
coalescent with recombination. By only considering those pairs of 
lineages that share 
genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
sequential Markovian structure along the chromosome the SMC is best known for.\\

More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as described below.\\

@@

Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
of a diploid population of constant size. The newly formed lineage acquires the 
union of both ordered intervals.
Although coalescence is reciprocal, here we'll define a strict total order 
(see \ref{par:liks}) on 
all lineages in $X(t)$ based on their leftmost starting points. 
For any such strict total order on $X_l(t)$ and at any point in the process 
the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see next paragraph). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
in case of a coalescence or recombination respectively.
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, an edge is registered along the full 
extent of both involved lineages. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

These unary nodes ensure the one to one correspondence between an ARG and its 
representation as a sequence of local trees (see fig.\ref{fig:smc-unary}). 
Indeed, the presence of these unary nodes allow us to completely reconstruct 
the state of all extant lineages through time from the local trees.

@@

As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes ensure these two 
representations contain the same information.
The haplotypes that make up the lineages coalescing 
at the root of each marginal tree are highlighted at the top. Ancestral 
material is coloured based on the samples that inherit from it.
Note how the presence of the unary nodes 
modifies the left-to-right logic of single-prune-regrafts induced by 
recombination under the SMC (blue dotted arrows).
Even in the absence of the blue recombination nodes, 
the unary coalescing nodes allow us to reconstruct the state of all lineages 
up to the exact time of the recombination event.
}
\label{fig:smc-unary}
\end{figure}

\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
Recording unary nodes has interesting consequences on the left-to-right 
description of the SMC. Looking at a single marginal tree, the presence of 
unary nodes are telltale signs of future or past coalescence events (in a 
left-to-right sense).\\

>>>>

Formulated backwards-in-time, the SMC \citep{mcvean_approximating_2005} only requires a 
simple modification to the 
coalescent with recombination. By only considering those pairs of 
lineages that share 
genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
sequential Markovian structure along the chromosome.\\

More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by a 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as outlined below.\\

@@

Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
of a diploid population of constant size. The newly formed lineage acquires the 
union of both ordered intervals.
Although coalescence is reciprocal, here we'll define a strict total order 
(see \ref{par:liks}) on 
all lineages in $X(t)$ based on their leftmost starting points. 
For any such strict total order on $X(t)$ and at any point in the process 
the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see \ref{par:liks). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
in case of a coalescence or recombination respectively.
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages affected by a recombination event. 
Coalescence between two lineages without perfectly 
overlapping intervals will similarly result in locally 
unary nodes (see \ref{fig:smc-unary}).These unary nodes ensure the one 
to one correspondence between an ARG and its 
representation as a sequence of local trees. 
Indeed, the presence of these unary nodes allow us to completely reconstruct 
the state of all extant lineages through time from the local trees.\\

@@

As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return the full this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes in the local trees 
ensure these two representations contain the same information. 
The ARG is augmented with haplotype information, coloured according to the 
samples that inherit from it. 
Note how the presence of a unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. It therefore modifies the 
left-to-right logic of the SMC whereby the floating lineage coalesces randomly
with a the remaining portion of the local tree. 
}
\label{fig:smc-unary}
\end{figure}

%\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
%Recording unary nodes has interesting consequences on the left-to-right 
%description of the SMC. Looking at a single marginal tree, the presence of 
%unary nodes are telltale signs of future or past coalescence events (in a 
%left-to-right sense).\\
